Solar wind predictions for the Parker Solar Probe orbit

### 2017-10-27 paperMVVB_referee/referee_report_responses_final.txt

Cover letter to the editor

Reference: AA/2017/31831

Reply to the reviewer comments on "Solar wind predictions for the Parker Solar Probe orbit - Near-Sun extrapolations derived from an empirical solar wind model based on Helios and OMNI observations", submitted by M. S. Venzmer and V. Bothmer to Astronomy and Astrophysics, 25 August 2017.

We are grateful to the referee for the thorough review of the manuscript and for the very helpful comments which we have carefully taken into account. All modifications appear as bold text in the revised version of the manuscript.

Major changes to the manuscript:
- We have adressed the general comments and the specific points made by the referee. We have added explanations and phrased some points more clearly.
- A new figure has been added, providing the predicted solar wind distributions for the first and the first closest perihelion.
- The discussion and summary sections have been combined. The equations have been moved from the summary to section 5.
- The tables and table captions have been improved for clarification.
- We found mismatching density values and corrected a typo in our density calculation. The error affected the densities for the first and nearest perihelia and the curves for the densities in figures 12 and 13. Corrections have been made to the abstract, section 6 and summary section: changed values 4015 to 214 and 9733 to 2951 cm-3.
- We updated Voyager 1’s distance in the introduction and the new sunspot numbers in figure 10.
- We have added a referee acknowledgment.
- We considered the suggestions concerning the English and further rephrased a few sentences for better readability.

#####

Reply to the referee

This paper discusses a method of using OMNI and Helios solar wind data to estimate the solar wind parameters that might be expected during the upcoming Parker Solar Probe mission. It includes an estimate of how these parameters might vary with solar cycle, though this is a relatively small correction, and the likely sunspot number during the mission is anyway highly uncertain. The analysis appears to be thorough, retaining information on the log-normal distribution of the parameters that is used to produce "model" representations of the distributions and their variation during the solar cycle. However, in the end, this information isn’t used in the predictions at PSP, where only median values are considered. It’s also not clear whether this type of analysis has any advantage over say a simple-minded approach where you use the radial dependence of averaged Helios values to propagate averaged 1 AU values, perhaps taken at different times in the solar cycle, down to the PSP orbit. My suspicion is that the results might not be all that different, but the authors might like to address this point. The extrapolated values do appear to be fairly consistent with previous estimates of near Sun parameters, and deviations, e.g. in the speed profile can be plausibly accounted for, so the results appear to be reasonable.

- We have tried answering the comments made through the specific comments below and by adding a new Figure providing the predicted solar wind distributions for the first overall, and the first “closest” perihelion. We have also explained that the results would not be that different but what our motivation was (see specific comments below).

@@

Some steps in the procedure could be explained more clearly in the text, as noted below, and the tables are rather cryptic since the meaning of the various parameters is not always clear. If they are fit parameters, it would be helpful to state the function that is being fitted in the table. There are conclusions and summary sections that both, yet again, go through the analysis procedure, which is unnecessary. They could be combined. There are also useful functions that give the radial dependences of the parameters that are only shown in the summary section. They could be included in the main results section.

- We have carefully incorporated the specific comments raised below and provided more explanations. The Tables and captions have been improved for clarification. The discussion and summary sections have been combined. The functions have been moved as suggested.

@@

The English is reasonable overall, but there are a number of cases where a long phrase is used as an adjective. Suggestions have been made to rephrase these sentences.

- We considered the suggestions and further rephrased a few sentences for better readability.

@@

Page 2, Results, line 5: Suggest: Since the modeled velocity and temperature values below about 20 Rs appear overestimated in comparison with existing observations, this suggests that PSP will directly measure solar wind acceleration and heating processes below 20 Rs as planned.

- We have modified the text accordingly.

Introduction, line 8: “Until today” is unnecessary – suggest removing. SOHO is also at L1 and returns solar wind observations from CELIAS. But maybe it’s not necessary to list the spacecraft, especially as other spacecraft not listed contributed earlier observations to OMNI.

- Removed as requested. We added a reference to the SOHO measurements, because we wanted to mention the current spacecraft in operation around L1 providing solar wind data.

@@

- Changed as suggested.

Page 6, Table 1:

Suggest writing the units together with the parameters in the first column, e.g. “Magnetic field (nT)”, rather than in a table footnote.

- Modified as suggested, also for all other Tables of the paper.

@@

I don’t really understand what the numbers in brackets (“estimated standard deviations”) represent. Why are they much larger than the parameter values and whole numbers?

- The explanation for the number in brackets has been added to the Table caption, as well as to all other Tables: The numbers in parentheses are the estimated standard deviations of the fit parameters referred to the corresponding last digits of the quoted value.

@@

How are the mean absolute errors and sums of absolute errors defined?

- The mean absolute errors are unnecessary in this context and are not referred to in the paper. So we removed them from the Table. The following text has been added to the caption, as well as it is mentioned in the text: For each parameter the sum of absolute residuals between data and fit (in percentage of the distribution area) is also listed.

@@

Page 7, Figure 1 caption, last sentence, suggest: show zoomed-in views of the high value tails of the distributions.

- Changed as suggested, also in caption of Figure 2.

Line 5 below caption: Supporting reference or other evidence that the extreme values are associated with “CME events” (ICMEs?)?

- We have clarified this aspect by adding: “The abnormally high parameter values in the distribution functions can be attributed to shock/CME events in agreement with the results of the OMNI solar wind investigations by Richardson and Cane (2012).”

@@

Line 7: Is this “overestimation” because [I]CMEs have abnormally low, not high, temperatures?

- Indeed this appears to be the case. We have picked this up and have also added a reference.

@@

Page 10, Table 2: Similar comments to those for Table 1 apply.

- Has been changed, see reply to Table 1.

Three lines below table and following: I find this step difficult to follow. It seems that the sunspot numbers are shifted according to the lags, and there is some fit to the data which is then used to produce a model of the particular parameter as a function of SSN, but the text doesn't explain the motivation (apparently to produce a "model" distribution as a function of SSN) procedure very clearly.

- Explanatory text has been added. Small changes added to section 3 too:
  - sentence with equation (9) clarified
  - second sentence: abbreviated 'sunspot number'
  - switched second and third paragraph
  - added a sentence to now second paragraph

@@

Below equation 8: I don't understand what "with the implementation of these relations into the lognormal function" means exactly. How are the fit coefficients obtained?

- The text has been modified to explain this point more clearly.

@@

Page 13, Table 3: Give the equation for the fit, to place the parameters in context.

- The reference to the fit equation is provided now, as is the case now for all other Table captions.

@@

Third paragraph below table, line 2: “See Figure 7”: Why not explain Figure 7 more fully here, then the coefficients d med, davg, etc. obtained from the fits will make more sense?

- Figure 7 has been introduced properly and the second reference moved to the end of this paragraph. A reference has been added to the figure caption.

@@

Last paragraph of page: “As anticipated…” - especially since the same data are used! Also give the results in the text to make the comparison easier with the values given in this paragraph.

- We removed “As anticipated” since it is obvious and we added the exponents we obtained.

@@

Page 14, last paragraph, line 6: The curves are still very close together after the crossing, so is this an important feature?

- We do not consider this aspect as a special feature but have clarified the crossing aspect in the text, see the answer to the next comment.

@@

Line 8: But this distance (0.082 AU) is beyond the range shown in the figure, and the data, so is it reliable? I don't understand the comment "therefore the extrapolation of the magnetic field and temperature distribution frequencies to the PSP orbit by applying lognormal functions is limited. What does “limited” mean? Perhaps the comment arises because the mean is larger than the median for a log normal distribution, as noted several pages earlier, and so this crossing indicates that the distribution is no longer log normal. Maybe a few more words of explanation would help the reader understand this point, and the significance of the crossing point.

- These distances are calculated from both fit curves. The questions have been addressed and clarified in the text, including the significance of the crossing point: “Thus, below that distance the frequency distribution cannot well be described anymore by a lognormal function, because the mean of a lognormal function has to be larger than its median… the crossing indicates that the parameter’s distribution is not anymore of a lognormal shape thereafter.” …  “The crossing points limit the regions where the distribution’s shapes can still be considered lognormal.”

@@

Line 10: How are the identical exponents chosen? E.g., was emed always used, or eavg, or some other value? This sentence reads as though the values were the same for all four parameters, but presumably it really means that a different exponent was used for each parameter.

- We have reworded the sentence for clarification, including a remark in the following paragraph.

@@

Page 16, line below figure: Not sure what "established" means here. Remove?

- “Established” has been removed.

Line 3: What parameter is used for the power law fits? Presumably it's the radial distance, but this could be stated specifically.

- The parameter is the radial distance and has been stated specifically now. A reference to the equation has been added.

@@

Page 17, figure caption: solar wind parameter power exponents for the dependence on radial distance

- Changed as suggested.

Last sentence of paragraph, suggest: Since this separation is highly variable, and requires significant effort to calculate for an extended time series, we have ignored this aspect in this analysis.

- Modified as suggested and added 'and' between references 'Bruno et al. (1986) and Balogh et al. (1999)'.

@@

Section 6, line 2, suggest: predictions of the SSN during the mission are incorporated into the empirical solar wind model derived in the previous sections

- Modified as suggested.

Next paragraph, last sentence, suggest: the perihelion distance to be reduced to a minimum of 9.86 Rs. Reverse figures 10 and 11, since figure 11 is referenced first, or just remove the mention of Figure 11 here, which isn't necessary; the statement can be understood without looking at the figure.

- Modified as suggested and switched order of figures 10 and 11. The time when the closest perihelion has been added to the first figure reference.

@@

Last line of page: How are the ranges of median values for different SSN shown in the figure? I'm guessing that it's the thickness of the shaded regions, but this could be stated specifically to guide the reader in interpreting the figure.

- A remark about this has been added, also to the figure caption.

@@

Figure 10 caption, line 1: The models obtained from Helios and OMNI measurements are …

- Modified as suggested and references to Sheeley 1997 and Wang 2000 have been ordered by year.

@@

Line 3: But do the extreme values really occur for min and max SSN? E.g. the highest speeds could be at intermediate SSN? How can a time lag be “noted” in this figure which there is no obvious time dependence?

- To avoid misunderstandings, we have rephrased the wording of the caption to: “The models obtained from Helios and OMNI measurements are extrapolated to the PSP region for SSN values between solar minimum and maximum, i.e., 0 ≤ ssn ≤ 200.” 
- Further on, the remark about the time lag has been removed from the figure caption. For clarification, a column for the time lag has been added to Table 2.

@@

Line 5, suggest: mentioned changing shape of the frequency distribution with heliocentric distance

- Is now as suggested.

Line 6: But the largest difference appears to be at ~1 AU, so isn't there also a problem with the dipole/quadrapole model matching the observed fields at 1 AU?

- This is correct and we have added an explaining sentence to the text: “It should be noted, that one cannot straightforward compare the absolute values of our study with the values obtained by Banaszkiewicz et al. because the DQCS model assumes solar wind originating from coronal holes at higher heliographic latitudes only, neglecting the slow solar wind belt.”

@@

Line 7: “Whereas”: Combine the sentences; whereas isn’t used at the beginning of a sentence. Since the peak speeds are on the declining phase of the cycle, wouldn't the peak predicted speeds occur for some intermediate sunspot number, not at SSN=200?

- The sentences have been combined. The comment is correct, it is a shifted SSN. We added the value of the time lag for each solar wind parameter.

@@

Last line: New sentence at “Thus”.

- Has been modified.

Page 19, line 6: What is the relevance of the comment on assuming a helium abundance? It's not clear how it relates to the comparison of the density profiles.

- The statement about the helium abundance has been reworded.

@@

Line 10: overestimates

- Corrected.

Paragraph 2, line 2: There is an abrupt change of topic here, and it's not immediately obvious why short term SSN predictions are relevant. There could be an introductory sentence saying that since the solar wind predictions depend on SSN, they may be refined using predictions of the SSN during the PSP mission.

- Introductory sentences have been incorporated, addressing short-term SSN predictions relevance as well as an additional argument on the prediction for the first perihelion.

@@

Line 5: Why should the last cycle be used rather than say a previous one? Is this based on expectations that the next cycle will also be small?

- Correct, we explained that we used the assumptions of Hathaway and Upton 2016.

@@

Line 8, suggest: the SSN during the closest perihelia, which will commence at the end of 2024 at the likely maximum phase of cycle 25, cannot be predicted at this time. Suggest moving this up to before "For the prediction of the next solar cycle..."

- Has been modified and moved as suggested.

Paragraph 3: How are the error bands in Figures 12 and 13 calculated? E.g., do they use different assumptions for the predicted SSN?

- We have explained now which SSN the predictions are based on and how the error bands have been calculated.

@@

Page 21, line 4, suggest: au, deriving power law fit functions that were used

- Modified as suggested.

Line 6, suggest: -2025, encompassing the prime mission up to the closest approach of 9.86 Rs.

- Modified as suggested. Replaced solar radii with Rs in discussion.

@@

Page 22, paragraph 2, line 5, suggest: that due to solar wind drag, the speeds

- Modified as suggested.

Summary, paragraph 2: It's not really necessary to go though the procedure yet again; there is already a synopsis in the conclusions. Is it necessary to have both a conclusions and a summary section? They could be combined.

- We have combined discussion and summary sections by integrating the first paragraph of the summary into the beginning of the discussion section and have removed the second description of the procedure. The section is now “Discussion and Summary”.

@@

Equation 12 and following equations: The formulas presented here have not been shown above, but are much easier to interpret than the results above. I'd suggest moving them into the results section above.

- We have moved the equations to the end of section 5 and added the formulas for the corresponding averages. The list statement about slow/fast wind balance has been implemented into the list statement about the velocity dependencies.

@@

Page 23, last but one paragraph: Again, there are points here that have already been said twice above.

- We have removed the paragraph.

#####

Additional changes applied:

- We found mismatching density values and corrected a typo in our density calculation. The error affected the densities for the first and nearest perihelia and the curves for the densities in figures 12 and 13. Corrections have been made to the abstract, section 6 and summary section: changed values 4015 to 214 and 9733 to 2951 cm-3.
- We have removed a duplicate 'during' in the caption of figure 13 and have replaced “nearest” with “first closest”.
- We changed km/s to km s-1 in all figures.
- We recalculated the error bands in figures 12 and 13. They changed slightly due to a previously not considered uncertainty in the error propagation.
- We changed the color in figure 14.
- We also added error bands for the velocity median values in figures 12 and 13.
- We updated the SSN and its predictions in figure 10 and noted in the caption which SIDC predictionmethod is plotted.
- Throughout the manuscript we corrected the appearance order of some phrases with their abbreviations. We further removed some abbreviations from the acknowledgments.
- We removed the article in the first sentence of figures 2 and 3.
- Introduction: We updated Voyager 1’s distance.
- We have added references to figure 7.
- We have adopted the formatting style of the A&A guidelines for some remaining elements (e.g., correct dash symbol).
- We have added a referee acknowledgment.

### 2017-11-05 paperMVVB/sections/abstract.tex
removed bold text

\date{received date; accepted date}
%First draft 9 August 2016; first changes 21 June 2017; second changes 17 July 2017; third changes 17 August 2017; fourth changes 24 August 2017; submitted 25 August 2017

\abstract
{The Parker Solar Probe (PSP) (formerly Solar Probe Plus) mission will be humanity’s first in situ exploration of the solar corona with closest perihelia at \num{9.86}~solar radii (\si{\Rs}) distance to the Sun. It will help answer hitherto unresolved questions on the heating of the solar corona and the source and acceleration of the solar wind and solar energetic particles. The scope of this study is to model the solar wind environment for PSP’s unprecedented distances during its prime mission phase during the years \numrange{2018}{2025}. The study is performed within the project Coronagraphic German And US Solar~Probe Survey (CGAUSS) which is the German contribution to the PSP mission as part of the Wide field Imager for Solar PRobe (WISPR).}	%context
{We present an empirical solar wind model for the inner heliosphere which is derived from OMNI and Helios data. The German-US space probes Helios~1 and Helios~2 flew in the 1970s and observed solar wind in the ecliptic within heliocentric distances of \SIrange{0.29}{0.98}{\au}. The OMNI database consists of multi-spacecraft intercalibrated in situ data obtained near \SI{1}{\au} over more than five solar cycles. The international sunspot number (SSN) and its predictions are used to derive dependencies of the major solar wind parameters on solar activity and to forecast their properties for the PSP mission.}	%aims
{The frequency distributions for the solar wind key parameters magnetic field strength, proton velocity, density and temperature are represented by lognormal functions. In addition, we consider the velocity distribution’s bi-componental shape, consisting of a slower and a faster part. Functional relations to solar activity are compiled with use of the OMNI data by correlating and fitting the frequency distributions with the SSN. Further, based on the combined data set from both Helios probes, the parameters’ frequency distributions are fitted with respect to solar distance to obtain power law dependencies. Thus an empirical solar wind model for the inner heliosphere confined to the ecliptic region is derived, accounting for solar activity and for solar distance through adequate shifts of the lognormal distributions. Finally, the inclusion of SSN predictions and the extrapolation to PSP’s perihelion enables us to estimate the solar wind environment for PSP’s planned trajectory during its mission duration.}	%methods
{The CGAUSS empirical solar wind model for PSP yields dependencies of the solar wind parameters on solar activity and radial distance. The estimated solar wind median values for PSP’s first perihelion in 2018 at a solar distance of \SI{0.16}{\au} are \SI{87}{\nT}, \SI{340}{\km\per\s}, \SI{214}{\per\cm\cubed} and \SI{503000}{\K}. The estimates for PSP’s closest perihelia, beginning in 2024 at \SI{0.046}{\au} (\SI{9.86}{\Rs}), are \SI{943}{\nT}, \SI{290}{\km\per\s}, \SI{2951}{\per\cm\cubed} and \SI{1930000}{\K}. \textbf{Since the modeled velocity and temperature values below about \SI{20}{\Rs} appear overestimated in comparison with existing observations, this suggests that PSP will directly measure solar wind acceleration and heating processes below \SI{20}{\Rs} as planned. } }	%results
{}	%conclusions

>>>>

\date{Received 25 August 2017; accepted date}
%First draft 9 August 2016; first changes 21 June 2017; second changes 17 July 2017; third changes 17 August 2017; fourth changes 24 August 2017; submitted 25 August 2017

\abstract
{The Parker Solar Probe (PSP) (formerly Solar Probe Plus) mission will be humanity’s first in situ exploration of the solar corona with closest perihelia at \num{9.86}~solar radii (\si{\Rs}) distance to the Sun. It will help answer hitherto unresolved questions on the heating of the solar corona and the source and acceleration of the solar wind and solar energetic particles. The scope of this study is to model the solar wind environment for PSP’s unprecedented distances in its prime mission phase during the years \numrange{2018}{2025}. The study is performed within the project Coronagraphic German And US SolarProbePlus Survey (CGAUSS) which is the German contribution to the PSP mission as part of the Wide-field Imager for Solar PRobe (WISPR).}	%context
{We present an empirical solar wind model for the inner heliosphere which is derived from OMNI and Helios data. The German-US space probes Helios~1 and Helios~2 flew in the 1970s and observed solar wind in the ecliptic within heliocentric distances of \SIrange{0.29}{0.98}{\au}. The OMNI database consists of multi-spacecraft intercalibrated in situ data obtained near \SI{1}{\au} over more than five solar cycles. The international sunspot number (SSN) and its predictions are used to derive dependencies of the major solar wind parameters on solar activity and to forecast their properties for the PSP mission.}	%aims
{The frequency distributions for the solar wind key parameters magnetic field strength, proton velocity, density and temperature are represented by lognormal functions. In addition, we consider the velocity distribution’s bi-componental shape, consisting of a slower and a faster part. Functional relations to solar activity are compiled with use of the OMNI data by correlating and fitting the frequency distributions with the SSN. Further, based on the combined data set from both Helios probes, the parameters’ frequency distributions are fitted with respect to solar distance to obtain power law dependencies. Thus an empirical solar wind model for the inner heliosphere confined to the ecliptic region is derived, accounting for solar activity and for solar distance through adequate shifts of the lognormal distributions. Finally, the inclusion of SSN predictions and the extrapolation down to PSP’s perihelion region enables us to estimate the solar wind environment for PSP’s planned trajectory during its mission duration.}	%methods
{The CGAUSS empirical solar wind model for PSP yields dependencies on solar activity and solar distance for the solar wind parameters' frequency distributions. The estimated solar wind median values for PSP’s first perihelion in 2018 at a solar distance of \SI{0.16}{\au} are \SI{87}{\nT}, \SI{340}{\km\per\s}, \SI{214}{\per\cm\cubed} and \SI{503000}{\K}. The estimates for PSP’s first closest perihelion, occuring in 2024 at \SI{0.046}{\au} (\SI{9.86}{\Rs}), are \SI{943}{\nT}, \SI{290}{\km\per\s}, \SI{2951}{\per\cm\cubed} and \SI{1930000}{\K}. Since the modeled velocity and temperature values below about \SI{20}{\Rs} appear overestimated in comparison with existing observations, this suggests that PSP will directly measure solar wind acceleration and heating processes below \SI{20}{\Rs} as planned.}	%results
{}	%conclusions

### 2017-11-05 paperMVVB/sections/acknowledgments.tex
removed bold text

\begin{acknowledgements}
	The authors acknowledge support of the Coronagraphic German and US SolarProbePlus Survey (CGAUSS) project for WISPR by the German Aerospace Center (DLR) under grant 50~OL~1601 as national contribution to the Parker Solar Probe mission. The authors thank the Helios and OMNI PIs/teams for creating and making available the solar wind in situ data. The Helios and the OMNI data are supplied by the NASA Space Science Data Coordinated Archive (NSSDCA) and the Space Physics Data Facility (SPDF) at NASA's Goddard Space Flight Center (GSFC). Additional thanks for maintaining and providing the international sunspot number series goes to the World Data Center -- Sunspot Index and Long-term Solar Observations (WDC-SILSO) at the Solar Influences Data Analysis Center (SIDC), Royal Observatory of Belgium (ROB). The PSP SPICE kernel was kindly provided by Angelos~Vourlidas. \textbf{The authors thank the referee for the careful review of this manuscript and helpful comments and suggestions.}
%	We thank the referees for /very important comments and suggestions. /careful review of this paper.
%	facilities/persons which made available the Helios and OMNI data\\
% 	``Ian Richardson and Hilary Cane for making their Interplanetary Coronal Mass Ejection list readily available''\\
% 	``Joseph King and Natalia Papitashvili for creating the combined OMNI data set''\\
% 	``We thank the reviewers for the careful review of this paper''\\
% 	The Authors extend their thanks to the referee for important comments and suggestions.\\
% 	We would like to thank XX for reading drafts of this manuscript...\\
% 	We would like to thank the Helios team and NASA's SPDF for supplying the plasma and magnetic field data (url)\\
\end{acknowledgements}

>>>>

\begin{acknowledgements}
	The authors acknowledge support of the Coronagraphic German and US SolarProbePlus Survey (CGAUSS) project for WISPR by the German Aerospace Center (DLR) under grant 50~OL~1601 as national contribution to the Parker Solar Probe mission. The authors thank the Helios and OMNI PIs/teams for creating and making available the solar wind in situ data. The Helios and the OMNI data are supplied by the NASA Space Science Data Coordinated Archive and the Space Physics Data Facility at NASA's Goddard Space Flight Center. Additional thanks for maintaining and providing the international sunspot number series goes to the World Data Center -- Sunspot Index and Long-term Solar Observations at the Solar Influences Data Analysis Center, Royal Observatory of Belgium. The PSP SPICE kernel was kindly provided by Angelos~Vourlidas. The authors thank the referee for the careful review of this manuscript and helpful comments and suggestions.
%	We thank the referees for /very important comments and suggestions. /careful review of this paper.
%	facilities/persons which made available the Helios and OMNI data\\
% 	``Ian Richardson and Hilary Cane for making their Interplanetary Coronal Mass Ejection list readily available''\\
% 	``Joseph King and Natalia Papitashvili for creating the combined OMNI data set''\\
% 	``We thank the reviewers for the careful review of this paper''\\
% 	The Authors extend their thanks to the referee for important comments and suggestions.\\
% 	We would like to thank XX for reading drafts of this manuscript...\\
% 	We would like to thank the Helios team and NASA's SPDF for supplying the plasma and magnetic field data (url)\\
\end{acknowledgements}
